---
title: "How to find tissue samples, using R and iDigBio"
output:
  html_document:
    code_folding: show
    df_print: kable
---

Please see **Use Case: [Find tissue samples](https://biodiversity-specimen-data.github.io/specimen-data-use-case/use-case/find-tissue-samples)** for context.

```{r message=FALSE}
#load core libraries; you will need to install these packages if you have not already
library(ridigbio)
library(tidyverse)

#load library for making nice HTML output via RMarkdown
library(kableExtra)
```

## Write a query to search for specimen records

First, let's find all the specimen records for the species you are interested in. Do this using the `idig_search` function from the ridigbio package. You can learn more about this function from the [iDigBio API documentation](https://github.com/iDigBio/idigbio-search-api/wiki) and [ridigbio documentation](https://cran.r-project.org/web/packages/ridigbio/ridigbio.pdf). In this example, we want to search for specimens identified as being in one of the following genera: _Manis_, _Rhinolophus_, or _Paguma_.

```{r}
#edit the fields (e.g. "genus") and values (e.g. "manis") in `list()` to adjust your query;
#edit the fields (e.g. "uuid") in `fields` to adjust the columns returned in your results;
records <- idig_search(rq = list(genus = c("manis","rhinolophus","paguma")),
                       fields = c("uuid", "recordset", "institutioncode",
                                  "genus", "scientificname",
                                  "country", "data.dwc:year",
                                  "data.dwc:collectionCode", "catalognumber",
                                  "data.dwc:preparations"))
```

The query above returned `r as.character(nrow(records))` records from the iDigBio Portal. Here is a preview of what our results look like:

```{r echo = FALSE, results = 'asis'}
knitr::kable(head(records)) %>% 
    kable_styling(bootstrap_options = 
                         c("striped", "hover", "condensed")) %>% 
  scroll_box(height = "300px")
```

## Filter specimen records by preparation type

These results include _all_ specimen records from the genera of interest, but now we would like to narrow our search to include only specimens that may have cryo-preserved tissue available. We are looking in the [dwc:preparations](https://dwc.tdwg.org/terms/#dwc:preparations) field for this information, but there isn't a controlled vocabulary in use and so the data is not very standard. We can summarize the distinct values present in this field for all `r as.character(nrow(records))` records:

``` {r}
#list distinct values for the preparation field
prepSummary <- records %>% 
  group_by(`data.dwc:preparations`) %>% 
  tally()

knitr::kable(prepSummary) %>% 
    kable_styling(bootstrap_options = 
                         c("striped", "hover", "condensed"),
                  full_width = FALSE) %>% 
  scroll_box(height = "400px")
```

### Select values to filter on

Based on the distinct values above, it seems like we can use the search terms "froze," "freeze," and "tissue" to narrow our results. We do this in the code below.

``` {r}
#edit the fields (e.g. "genus") and values (e.g. "manis") in `list()` to adjust your query;
recordsFiltered <- records %>% 
  mutate(`data.dwc:preparations` = str_to_lower(`data.dwc:preparations`)) %>% 
  filter(grepl('freeze|froze|tissue', `data.dwc:preparations`))
```

### Apply filter

Now we have `r as.character(nrow(recordsFiltered))` records, all of which are displayed in the table here:

```{r echo = FALSE, results = 'asis'}
knitr::kable(recordsFiltered) %>% 
    kable_styling(bootstrap_options = 
                         c("striped", "hover", "condensed")) %>% 
  scroll_box(height = "600px")

#if you have this code open in R, you can uncomment the line below to save recordsFiltered as a csv file
#write_csv(recordsFiltered, "recordsFiltered.csv")
```

We can view the preparation distinct values present in our subset of data to see if there are any other values we might want to filter on:

``` {r echo = FALSE, results = 'asis'}
#list distinct values for the preparation field in recordsFiltered
recordsFiltered %>% 
  group_by(`data.dwc:preparations`) %>% 
  tally() %>% 
  knitr::kable() %>% 
    kable_styling(bootstrap_options = 
                         c("striped", "hover", "condensed"),
                  full_width = FALSE) %>% 
  scroll_box(height = "400px")
```

## Contact institutions with specimens of interest

These data come from from multiple institutions, and if we wanted to examine the physical specimens or inquire about getting tissue samples, we would need to contact each institution. We can figure out who to contact using our results, as shown by another [use case](https://biodiversity-specimen-data.github.io/specimen-data-use-case/use-case/identify-data-contacts) and [solution](https://biodiversity-specimen-data.github.io/specimen-data-use-case/solution/identify-data-contacts_ridigbio).